---
title: "Motif analysis with Signac for sc-ATAC-Seq Analysis in 10X Genomics data"
author: "Antonia Chroni for SJCRH DNB_BINF_Core"
papersize: a4
fontsize: 11pt
links-as-notes: true
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    df_print: paged
    code_folding: hide
    toc_depth: 2
    highlight: tango
    number_sections: TRUE
  pdf_document:
    toc: TRUE
    highlight: tango
    number_sections: TRUE
    latex_engine: lualatex
    keep_tex: FALSE
    fig_caption: yes
    fig_crop: no
    fig_height: 2
    fig_width: 3
    toc_depth: 2
always_allow_html: TRUE
urlcolor: blue
linkcolor: black
citecolor: blue
geometry: margin=1in
header-includes: 
  - \usepackage{titling}
  - \usepackage{fancyhdr}
  - \usepackage{graphicx}
  - \usepackage{float}
params:
  data_dir_motif: '.'
  data_file_motif: '.'
  assay: '.'
  resolution_list: '.'
  genome: '.'
  min.cutoff_value: '.'
  max.cutoff_value: '.'
  top_n_value_motifs: '.'
  top_da_peak_names_motifs: '.'
  cell_type_name: '.'
  condition_value1: '.'
  condition_value2: '.'
  condition_value3: '.'
  root_dir: './'
  PROJECT_NAME: '.'
  PI_NAME: '.'
  TASK_ID: '.'
  PROJECT_LEAD_NAME: '.'
  DEPARTMENT: '.'
  LEAD_ANALYSTS: '.'
  GROUP_LEAD: '.'
  CONTACT_EMAIL: '.'
  PIPELINE: '.'
  START_DATE: '.'
  COMPLETION_DATE: '.'
---

```{r logo-file, echo=FALSE}
attach(params)
# Insert logo on the top of the html report 
logo_file <- file.path(root_dir, "figures", "img", "DNB-BINF-Core-logo.png")
htmltools::img(src = knitr::image_uri(logo_file), alt = "logo", style = "position:absolute; top:0; left:0; padding:0px; height:120px;")
detach(params)
```

\addtolength{\headheight}{2.0cm} 
\fancypagestyle{plain}{} 
\thispagestyle{fancy}
\fancyhead[L]{\includegraphics[height=120px]{`r logo_file`}}
\renewcommand{\headrulewidth}{0pt}

<style type="text/css">
:root {--DNB_BINF_Core_color: #00427B;}

h1.title {margin-top: 130px;
          margin-bottom: 25px;
          font-size: 36px;}

.nobullet li {list-style-type: none;}

.reporthead {font-size: 20px;}

body { /* Normal */
  font-size: 16px;
  font-style: Arial, Helvetica, sans-serif;}

h1 {color: var(--DNB_BINF_Core_color);
    font-size: 28px;
    margin-top: 50px;}

h2 {color: var(--DNB_BINF_Core_color);
    font-size: 20px;}

.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
  background-color: var(--DNB_BINF_Core_color);}
</style>

<a href="https://wiki.stjude.org/display/CAB">

</a>

\pagebreak

<div class="reporthead"><br/>
**PI: `r params$PI_NAME`**  
**Project: `r params$PROJECT_NAME`**  
Task: `r params$TASK_ID`  
Project Lead(s): `r params$PROJECT_LEAD_NAME`  
Department: `r params$DEPARTMENT`  

<br />  

DNB Bioinformatics Core Analysis Team: 
<br />  

>**Lead Analyst(s): `r params$LEAD_ANALYSTS`**  
>Group Lead: `r params$GROUP_LEAD`  
<br />
>**Contact E-mail:** `r params$CONTACT_EMAIL`  
>**DNB Bioinformatics Core Pipeline:** `r params$PIPELINE`  

Date started: `r params$START_DATE`  
Date completed:  `r params$COMPLETION_DATE`  
Report generated: `r format(Sys.time(), '%H:%M:%S %Z %m/%d/%Y')` \

Reviewed by: _____________________   Date: ____________ \
</div>
\pagebreak
  
# Information about this notebook

This notebook guides users in performing DNA sequence motif analysis in Signac. We will explore two complementary options for performing motif analysis: one by finding over-represented motifs in a set of differentially accessible peaks, one method performing differential motif activity analysis between groups of cells.

For more information, see [Motif analysis with Signac](https://stuartlab.org/signac/articles/motif_vignette).

## What are motifs?

- **A motif is a short, recurring DNA sequence pattern** that is statistically enriched in regulatory regions (like promoters or enhancers) and has biological meaning.
- Motifs can be found in DNA, RNA, and proteins.
- They are often involved in the regulation of gene expression, protein folding, and other biological processes.
- Motifs represent the binding preference of transcription factors (TFs). For example, a TF might recognize a specific 6–10 bp motif.

## What are transcription factors (TF)?

- **A transcription factor is a protein that binds to DNA at specific sites (often motifs) to regulate gene expression.**
- TFs are proteins that bind to specific DNA sequences (often called motifs) to regulate the transcription of genetic information from DNA to RNA.
- TFs are proteins that control the rate of transcription of genetic information from DNA to messenger RNA (mRNA).
- Can be either DNA- or RNA-binding proteins.
- Bind to specific sequences of DNA called promoter sequences.
- The binding to promoter sequences can either stimulate or repress transcription.
- Essential for regulating gene expression, which is the process by which genes are turned on and off.
- Can be activated by a variety of signals, including hormones, growth factors, and environmental cues.
- Can be regulated by post-transnational modifications, such as phosphorylation and acetylation.

Based on their structure or function, some of the most common types of TFs include:

- **Activators**: Stimulate transcription by binding to promoter sequences and recruiting RNA polymerase.
- **Repressors**: Repress transcription by binding to promoter sequences and preventing RNA polymerase from binding.
- **Coactivators**: Work with activators to stimulate transcription.
- **Corepressors**: Work with repressors to repress transcription.

## What research questions does Motif analysis help to answer?

**(1) Motif Discovery**

- What are the binding preferences of transcription factors (TFs)?
Identify sequence patterns that TFs recognize in DNA.

- How do transcription factors regulate gene expression?
Understand how TF binding influences transcriptional activation or repression.

- How does DNA sequence variation affect TF binding?
Explore the impact of SNPs or mutations on regulatory element function.

**(2) Motif Enrichment**

- Which genes are regulated by a particular transcription factor?
Link motif presence to target gene networks.

- How do motifs contribute to gene regulation across cell types or conditions?
Compare motif activity in different tissues, developmental stages, or disease states.

- What role do motifs play in disease development?
Investigate how altered motif accessibility or binding drives pathogenesis.

- Can transcription factors be targeted for therapeutic intervention?
Assess motif-driven regulatory pathways as potential drug targets.

## References

- [Carels, N. (2015). A History of Genomic Structures: The Big Picture. In: Bahadur, B., Venkat Rajam, M., Sahijram, L., Krishnamurthy, K. (eds) Plant Biology and Biotechnology. Springer, New Delhi.](https://doi.org/10.1007/978-81-322-2283-5_7)
- [Introduction to Motif Discovery and Transcription Factor Binding Site Analysis](https://www.youtube.com/watch?v=dWvkSFeQg9Q&t=2905s)


# Set up

```{r load-library, echo=TRUE}
attach(params)
suppressPackageStartupMessages({
  library(Signac)
  library(Seurat)
  library(DT)
  library(patchwork)
  library(tidyverse)
  library(JASPAR2020)
  library(TFBSTools)
  library(knitr)
  library(GenomeInfoDb) #translation between chromosome names
  library(BSgenome)
  library(motifmatchr)
  library(BSgenome.Mmusculus.UCSC.mm10)
  library(BSgenome.Hsapiens.UCSC.hg38)
  
  library(BSgenome.Mmusculus.UCSC.mm39)
  library(ggseqlogo)
  library(chromVAR)
}) 
```

# Directories and paths to file Inputs/Outputs

```{r set-dir-and-file-names, echo=TRUE}
module_analysis_dir <- file.path(root_dir, "analyses", "motif-footprint-analysis") 

# Input files
data_file <- file.path(data_dir_motif, data_file_motif)
gradient_palette_file <- file.path(root_dir, "figures", "palettes", "gradient_color_palette.tsv")

# Create plots directory
module_plots_dir <- file.path(module_analysis_dir , "plots")

plots_dir <- file.path(module_plots_dir, "01_motif_analysis") 
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)}

ChromVAR_plots_dir <- file.path(plots_dir, "01_ChromVAR_analysis") 
if (!dir.exists(ChromVAR_plots_dir)) {
  dir.create(ChromVAR_plots_dir)}

# Create results directory
module_results_dir <- file.path(module_analysis_dir , "results")
if (!dir.exists(module_results_dir)) {
  dir.create(module_results_dir)}

results_dir <- file.path(module_results_dir, "01_motif_analysis") 
if (!dir.exists(results_dir)) {
  dir.create(results_dir)}

# Source functions
source(paste0(module_analysis_dir, "/util/function-add-motif.R"))
source(paste0(module_analysis_dir, "/util/function-create-FeaturePlot.R"))
source(paste0(root_dir, "/figures/scripts/theme_plot.R"))
```

```{r echo=FALSE, warning=FALSE}
opts_chunk$set(fig.align='center',
               external=TRUE,
               echo=FALSE,
               warning=FALSE,
               fig.pos='H')
a4width <- 8.3
a4height <- 11.7
```

# Read seurat object

First, we will use the seurat object as generated from the pipeline in the `r data_dir_motif` directory. 

```{r read-object, fig.width = 6, fig.height = 5, fig.fullwidth = TRUE, echo=TRUE}
seurat_obj <- readRDS(data_file)
DefaultAssay(seurat_obj) <- assay

# Set identity classes to an existing column in meta data
Idents(seurat_obj) <- seurat_obj@meta.data[[cell_type_name]]

# View cell identities, get summary table
table(Idents(seurat_obj))

# Plot cell types
name <- paste0(plots_dir, "/", glue::glue("DimPlot-cell-types.png"), sep = "")
p1 <- print(DimPlot(seurat_obj, label = TRUE, pt.size = 0.1) + NoLegend())
ggsave(file = name, width = 6, height = 5, device = "png")
```

```{r read-palette, echo=TRUE}
gradient_palette_df <- readr::read_tsv(gradient_palette_file, guess_max = 100000, show_col_types = FALSE) 
```

# Adding motif information to the Seurat object

We will run the [AddMotifs()](https://stuartlab.org/signac/reference/addmotifs) function to add the DNA sequence motif information required for motif analyses.

To facilitate motif analysis in Signac, there is the Motif class to store all the required information, including a list of position weight matrices (PWMs) or position frequency matrices (PFMs) and a motif occurrence matrix. 

Here, the [AddMotifs()](https://stuartlab.org/signac/reference/addmotifs) function constructs a Motif object and adds it to the dataset, along with other information such as the base composition of each peak. A motif object can be added to any Seurat assay using the [SetAssayData()](https://satijalab.github.io/seurat-object/reference/AssayData.html) function. See the [object interaction vignette](https://stuartlab.org/signac/articles/data_structures) for more information.

```{r add-motif, echo=TRUE}
motifs_result <- add_motifs(seurat_obj = seurat_obj, assay = assay, genome = genome, pfm = pfm) 

# Extract the objects
seurat_obj <- motifs_result$seurat_obj
motif_obj <- motifs_result$motif_obj
```

# Finding overrepresented motifs

To identify potentially important cell-type-specific regulatory sequences, we can search for DNA motifs that are overrepresented in a set of peaks that are differentially accessible between cell types.

```{r da-peaks, echo=TRUE}
set.seed(1234)

#da_peaks <- FindMarkers(
da_peaks <- FindAllMarkers(object = seurat_obj,
                           assay = assay, 
                           #ident.1 = 'Pvalb',
                           #ident.2 = 'Sst',
                           only.pos = TRUE, # FALSE: include both up and down peaks
                           test.use = 'LR',
                           min.pct = 0.05,
                           latent.vars = 'nCount_peaks',
                           min.diff.pct = 0.1) # for stronger effect size filtering and to avoid weak differences
write_tsv(da_peaks, file = paste0(results_dir, "/", "da_peaks_all.tsv")) 

# Consider multiple testing correction
# We will use the p_val_adj instead of raw p_val for more robust filtering
top_da_peaks <- da_peaks %>%
  filter(p_val_adj < 0.05, pct.1 > 0.2) %>%
  arrange(p_val) %>%
  write_tsv(file.path(results_dir, "da_peaks_top.tsv"))
```

## Interactive Table of differentially accessible peaks by cell type

```{r da-peaks-interactive-table, echo=TRUE}
# Interactive Table
datatable(da_peaks, 
          options = list(pageLength = 5, 
                         autoWidth = TRUE, 
                         server = TRUE), 
          filter = "top")
```

### Optional: choosing a set of background peaks - Motif Enrichment

We will select the top `r top_da_peak_names_motifs` and we will plot the position weight matrices for the motifs, so we can visualize the different motif sequences.

```{r MotifPlot, fig.width = 20, fig.height = 10, fig.fullwidth = TRUE, echo=TRUE}
# Perform a hypergeometric test to test the probability of observing the motif at the given frequency by chance, comparing with a background set of peaks matched for GC content.
# Extract peak names (features) as a character vector
top_da_peak_names <- rownames(top_da_peaks)

# Take the first three features from top_da_peak_names
enriched.motifs <- FindMotifs(object = seurat_obj,
                              features = top_da_peak_names[1:top_da_peak_names_motifs])    #features = top_da_peak_names

write_tsv(enriched.motifs, file = paste0(results_dir, "/", "da_peaks_top_enriched_motifs.tsv"))

name <- paste0(plots_dir, "/", "MotifPlot_enriched_motifs_da_peaks_top.png", sep = "")
print(MotifPlot(object = seurat_obj, motifs = head(rownames(enriched.motifs))))
ggsave(file = name, width = 20, height = 10, device = "png")
```

### Interactive Table of Enriched Motifs

```{r enriched-motifs-interactive-table, echo=TRUE}
# Interactive Table
datatable(enriched.motifs, 
          options = list(pageLength = 5, 
                         autoWidth = TRUE, 
                         server = TRUE), 
          filter = "top")
```

# Computing motif activities per cell

We can compute per-cell motif activity scores using [chromVAR](https://greenleaflab.github.io/chromVAR/index.html). This allows us to visualize motif activities per cell, and also provides an alternative method of identifying differentially-active motifs between cell types. ChromVAR identifies motifs associated with variability in chromatin accessibility between cells. See the [chromVAR paper](https://www.nature.com/articles/nmeth.4401) for a complete description of the method.

```{r RunChromVAR, echo=TRUE}
seurat_obj <- RunChromVAR(object = seurat_obj, genome = genome)
DefaultAssay(seurat_obj) <- 'chromvar'
saveRDS(seurat_obj, file = paste0(results_dir, "/", "seurat_chromvar.rds"))  # Save object

# Sanity checks:Inspect the assay
seurat_obj[["chromvar"]]             # should show features = motifs, slot names: counts/data/scale.data
head(rownames(seurat_obj[["chromvar"]]))  # motif names
#summary(seurat_obj[["chromvar"]]@data[])   # distribution of deviation Z-scores
```

## Plotting motif activities across cell types

We will select the **top `r top_n_value_motifs`** differentially accessible peaks per cluster for plotting motif activities.

```{r top-motifs-select, echo=TRUE}
chromvar_mat <- seurat_obj[["chromvar"]]@data # Get the chromVAR deviation matrix
avg_scores <- rowMeans(chromvar_mat) # Compute average deviation Z-score per motif across all cells

# Sort and pick the top motifs
top_motifs <- sort(avg_scores, decreasing = TRUE)[1:top_n_value_motifs] 
print(top_motifs) # Display as a named vector
```

### Overall

```{r plot-RunChromVAR-overall, fig.width = 15, fig.height = 5, fig.fullwidth = TRUE, echo=TRUE}
plot_RunChromVAR_variable_overall <- create_featureplot_variable_v1(seurat_obj = seurat_obj, palette = gradient_palette_df, min.cutoff = min.cutoff_value, max.cutoff = max.cutoff_value, ChromVAR_plots_dir = ChromVAR_plots_dir, split_variable = NULL)
```

### Split by ID

```{r plot-RunChromVAR-ID, fig.width = 15, fig.height = 5, fig.fullwidth = TRUE, echo=TRUE}
plot_RunChromVAR_variable_ID <- create_featureplot_variable_v1(seurat_obj = seurat_obj, palette = gradient_palette_df, min.cutoff = min.cutoff_value, max.cutoff = max.cutoff_value, ChromVAR_plots_dir = ChromVAR_plots_dir, split_variable = "ID")
```

## Split by `r condition_value1`

We will group the motif activities by the **`r condition_value1`** metadata column. 

```{r plot-RunChromVAR-condition-value1, fig.width = 15, fig.height = 5, fig.fullwidth = TRUE, echo=TRUE}
plot_RunChromVAR_variable_value1 <- create_featureplot_variable_v1(seurat_obj = seurat_obj, palette = gradient_palette_df, min.cutoff = min.cutoff_value, max.cutoff = max.cutoff_value, ChromVAR_plots_dir = ChromVAR_plots_dir, split_variable = condition_value1) 
```

## Split by `r condition_value2`

We will group the motif activities by the **`r condition_value2`** metadata column. 

```{r plot-RunChromVAR-condition-value2, fig.width = 15, fig.height = 5, fig.fullwidth = TRUE, echo=TRUE}
if (!is.null(condition_value2)) {
  stopifnot(is.character(condition_value2), length(condition_value2) == 1)
  stopifnot(condition_value2 %in% colnames(seurat_obj@meta.data))
  plot_RunChromVAR_variable_value2 <- create_featureplot_variable_v1(seurat_obj = seurat_obj, palette = gradient_palette_df, min.cutoff = min.cutoff_value, max.cutoff = max.cutoff_value, ChromVAR_plots_dir = ChromVAR_plots_dir, split_variable = condition_value2)
}
```

## Split by `r condition_value3`

We will group the motif activities by the **`r condition_value3`** metadata column. 

```{r plot-RunChromVAR-condition-value3, fig.width = 15, fig.height = 5, fig.fullwidth = TRUE, echo=TRUE}
if (!is.null(condition_value3)) {
  stopifnot(is.character(condition_value3), length(condition_value3) == 1)
  stopifnot(condition_value3 %in% colnames(seurat_obj@meta.data))
  plot_RunChromVAR_variable_value3 <- create_featureplot_variable_v1(seurat_obj = seurat_obj, palette = gradient_palette_df, min.cutoff = min.cutoff_value, max.cutoff = max.cutoff_value, ChromVAR_plots_dir = ChromVAR_plots_dir, split_variable = condition_value3)
}
```

```{r echo=FALSE}
detach(params)
```

\pagebreak

# Session Info

```{r echo=FALSE}
sessionInfo()
```
