---
title: "Building trajectories with Monocle 3 for sc-ATAC-Seq Analysis in 10X Genomics data"
author: "Antonia Chroni for SJCRH DNB_BINF_Core"
papersize: a4
fontsize: 11pt
links-as-notes: true
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    df_print: paged
    code_folding: hide
    toc_depth: 2
    highlight: tango
    number_sections: TRUE
  pdf_document:
    toc: TRUE
    highlight: tango
    number_sections: TRUE
    latex_engine: lualatex
    keep_tex: FALSE
    fig_caption: yes
    fig_crop: no
    fig_height: 2
    fig_width: 3
    toc_depth: 2
always_allow_html: TRUE
urlcolor: blue
linkcolor: black
citecolor: blue
geometry: margin=1in
header-includes: 
  - \usepackage{titling}
  - \usepackage{fancyhdr}
  - \usepackage{graphicx}
  - \usepackage{float}
params:
  assay: '.'
  resolution_list: '.'
  lineage1_value: '.'
  lineage1_value_vector: '.'
  lineage2_value: '.'
  lineage2_value_vector: '.'
  lineage_value: '.'
  root_dir: './'
  PROJECT_NAME: '.'
  PI_NAME: '.'
  TASK_ID: '.'
  PROJECT_LEAD_NAME: '.'
  DEPARTMENT: '.'
  LEAD_ANALYSTS: '.'
  GROUP_LEAD: '.'
  CONTACT_EMAIL: '.'
  PIPELINE: '.'
  START_DATE: '.'
  COMPLETION_DATE: '.'
---

```{r logo-file, echo=FALSE}
attach(params)
# Insert logo on the top of the html report 
logo_file <- file.path(root_dir, "figures", "img", "DNB-BINF-Core-logo.png")
htmltools::img(src = knitr::image_uri(logo_file), alt = "logo", style = "position:absolute; top:0; left:0; padding:0px; height:120px;")
detach(params)
```

\addtolength{\headheight}{2.0cm} 
\fancypagestyle{plain}{} 
\thispagestyle{fancy}
\fancyhead[L]{\includegraphics[height=120px]{`r logo_file`}}
\renewcommand{\headrulewidth}{0pt}

<style type="text/css">
:root {--DNB_BINF_Core_color: #00427B;}

h1.title {margin-top: 130px;
          margin-bottom: 25px;
          font-size: 36px;}

.nobullet li {list-style-type: none;}

.reporthead {font-size: 20px;}

body { /* Normal */
  font-size: 16px;
  font-style: Arial, Helvetica, sans-serif;}

h1 {color: var(--DNB_BINF_Core_color);
    font-size: 28px;
    margin-top: 50px;}

h2 {color: var(--DNB_BINF_Core_color);
    font-size: 20px;}

.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
  background-color: var(--DNB_BINF_Core_color);}
</style>

<a href="https://wiki.stjude.org/display/CAB">

</a>

\pagebreak

<div class="reporthead"><br/>
**PI: `r params$PI_NAME`**  
**Project: `r params$PROJECT_NAME`**  
Task: `r params$TASK_ID`  
Project Lead(s): `r params$PROJECT_LEAD_NAME`  
Department: `r params$DEPARTMENT`  

<br />  

DNB Bioinformatics Core Analysis Team: 
<br />  

>**Lead Analyst(s): `r params$LEAD_ANALYSTS`**  
>Group Lead: `r params$GROUP_LEAD`  
<br />
>**Contact E-mail:** `r params$CONTACT_EMAIL`  
>**DNB Bioinformatics Core Pipeline:** `r params$PIPELINE`  

Date started: `r params$START_DATE`  
Date completed:  `r params$COMPLETION_DATE`  
Report generated: `r format(Sys.time(), '%H:%M:%S %Z %m/%d/%Y')` \

Reviewed by: _____________________   Date: ____________ \
</div>
\pagebreak
  
# Information about this notebook

This notebook guides users in inferring trajectories and pseudotime on single-cell ATAC-seq (scATAC-seq) data.

For more information, see [Building trajectories with Monocle 3](https://stuartlab.org/signac/articles/monocle) and [Monocle 3](https://cole-trapnell-lab.github.io/monocle3/).

# Set up

```{r load-library, echo=TRUE}
attach(params)
suppressPackageStartupMessages({
  library(Signac)
  library(Seurat)
  library(SeuratWrappers)
  library(monocle3)
  library(Matrix)
  library(tidyverse)
  library(patchwork)
  library(base)
  library(knitr)
  library(scooter)
}) 
```

# Directories and paths to file Inputs/Outputs

```{r set-dir-and-file-names, echo=TRUE}
module_analysis_dir <- file.path(root_dir, "analyses", "building-trajectories-monocle") 
analysis_dir <- file.path(root_dir, "analyses", "integration-with-scrna-seq-data") 
data_dir <- file.path(analysis_dir, "results", "01_integration_with_scrna_seq_data")

# Input files
data_file <- file.path(data_dir, "seurat_obj_reference.rds")

# Create plots directory
module_plots_dir <- file.path(module_analysis_dir , "plots")

plots_dir <- file.path(module_plots_dir, "01_building_trajectories_monocle") 
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)}

# Create results directory
module_results_dir <- file.path(module_analysis_dir , "results")
if (!dir.exists(module_results_dir)) {
  dir.create(module_results_dir)}

results_dir <- file.path(module_results_dir, "01_building_trajectories_monocle") 
if (!dir.exists(results_dir)) {
  dir.create(results_dir)}

# Source functions
source(paste0(module_analysis_dir, "/util/monocle3_updated.R"))
```

```{r echo=FALSE, warning=FALSE}
opts_chunk$set(fig.align='center',
               external=TRUE,
               echo=FALSE,
               warning=FALSE,
               fig.pos='H')
a4width <- 8.3
a4height <- 11.7
```

# Read seurat object

First, we will use the seurat object as generated from the pipeline in the `integration-with-scrna-seq-data` module. 

```{r read-object, fig.width = 6, fig.height = 5, fig.fullwidth = TRUE, echo=TRUE}
seurat_obj <- readRDS(data_file)
DefaultAssay(seurat_obj) <- assay

# Set the identity of your cells to the column with the desired resolution
use_res_col <- glue::glue("{assay}_snn_res.{resolution_list}")

# Set identity classes to an existing column in meta data
Idents(object = seurat_obj) <- use_res_col

# View cell identities, get summary table
#Idents(seurat_obj)
table(Idents(seurat_obj))


name <- paste0(plots_dir, "/", glue::glue("DimPlot-clusters.png"), sep = "")
print(DimPlot(seurat_obj, label = TRUE) + NoLegend())
ggsave(file = name, width = 6, height = 5, device = "png")
```

Assign each cluster to the most common cell type based on the previous predicted annotations from the `./integration-with-scrna-seq-data/01-integration-with-scrna-seq-data.Rmd` step.

```{r assign-celltypes, fig.width = 6, fig.height = 5, fig.fullwidth = TRUE, echo=TRUE}
for(i in levels(seurat_obj)) {
  cells_to_reid <- WhichCells(seurat_obj, idents = i)
  newid <- names(sort(table(seurat_obj$predicted.id[cells_to_reid]),decreasing=TRUE))[1]
  Idents(seurat_obj, cells = cells_to_reid) <- newid
}

seurat_obj$assigned_celltype <- Idents(seurat_obj)

name <- paste0(plots_dir, "/", glue::glue("DimPlot-assigned-celltype.png"), sep = "")
print(DimPlot(seurat_obj, label = TRUE))
ggsave(file = name, width = 6, height = 5, device = "png")
```

Next we can subset the different lineages and create a trajectory for each lineage. Another way to build the trajectories is to use the whole dataset and build separate pseudotime trajectories for the different cell partitions found by Monocle 3.

```{r subset-celltypes, echo=TRUE}
lineage1 <- seurat_obj[,  seurat_obj$assigned_celltype %in% lineage1_value_vector]
lineage2 <- seurat_obj[, seurat_obj$assigned_celltype %in% lineage2_value_vector]
```

# Building trajectories with Monocle 3

We can convert the Seurat object to a CellDataSet object using the as.cell_data_set() function from [SeuratWrappers](https://github.com/satijalab/seurat-wrappers) and build the trajectories using Monocle 3. We’ll do this separately for each lineage, but you could explore other strategies building a trajectory for all lineages together.

```{r trajectories-each-lineage, echo=TRUE}
# We need to add this helper function at the top of our script to mimic the Seurat v4 internal function that is missing in the current version:
AssociatedDimReducs <- function(object, assay = NULL) {
  assay <- assay %||% DefaultAssay(object)
  reduc.names <- names(Filter(function(x) {
    x@assay.used == assay
  }, object@reductions))
  return(reduc.names)
}

set.seed(2025)

lineage1.cds <- as.cell_data_set(lineage1)
lineage1.cds <- cluster_cells(cds = lineage1.cds, reduction_method = "UMAP")
lineage1.cds <- learn_graph(lineage1.cds, use_partition = TRUE)

lineage2.cds <- as.cell_data_set(lineage2)
lineage2.cds <- cluster_cells(cds = lineage2.cds, reduction_method = "UMAP")
lineage2.cds <- learn_graph(lineage2.cds, use_partition = TRUE)
```

To compute pseudotime estimates for each trajectory we need to decide what the start of each trajectory is. 

Monocle 3 includes an interactive function to select cells as the root nodes in the graph. This function will be launched if calling [order_cells()](https://rdrr.io/pkg/monocle3/man/order_cells.html) without specifying the root_cells parameter.

```{r plot-trajectories, fig.width = 6, fig.height = 5, fig.fullwidth = TRUE, echo=TRUE}
set.seed(2025)

# order cells
# Find barcodes for lineage1 cells
lineage1_cells <- colnames(lineage1.cds)[colData(lineage1.cds)$assigned_celltype == lineage1_value_vector]
# Use the first lineage1 cell as the root (or a vector of root cells)
lineage1.cds <- order_cells(lineage1.cds, reduction_method = "UMAP", root_cells = lineage1_cells[1])

# Find barcodes for lineage2 cells
lineage2_cells <- colnames(lineage2.cds)[colData(lineage2.cds)$assigned_celltype == lineage2_value_vector]
# Use the first lineage2 cell as the root (or a vector of root cells)
lineage2.cds <- order_cells(lineage2.cds, reduction_method = "UMAP", root_cells = lineage2_cells[1])


# Plot trajectories colored by pseudotime
name <- paste0(plots_dir, "/", glue::glue("Plot-trajectory-lineage1.png"), sep = "")
print(plot_cells(cds = lineage1.cds, color_cells_by = "pseudotime", show_trajectory_graph = TRUE))
ggsave(file = name, width = 6, height = 5, device = "png")

name <- paste0(plots_dir, "/", glue::glue("Plot-trajectory-lineage2.png"), sep = "")
print(plot_cells(cds = lineage2.cds, color_cells_by = "pseudotime", show_trajectory_graph = TRUE))
ggsave(file = name, width = 6, height = 5, device = "png")
```

# How to Programmatically Root Cells in a Trajectory Analysis

For exploratory trajectory analysis, it is often useful to root cells based on a specific cluster or cell type. In this example, we root the trajectory using cells from a cluster defined by the `predicted.id` column in the metadata, with the lineage of interest specified by the `r lineage_value` as defined in parameters.

For more information, see the [Monocle3 issue](https://github.com/cole-trapnell-lab/monocle3/issues/328).

```{r programmatically-root-cells, echo=TRUE}
cds <- as.cell_data_set(seurat_obj) # Convert your Seurat object to a Monocle3 cell_data_set.
cds <- cluster_cells(cds = cds, reduction_method = "UMAP") # Cluster cells and learn the principal graph using Monocle3.
cds <- learn_graph(cds, use_partition = TRUE) # to choose more than one node
#cds <- learn_graph(cds, use_partition = FALSE) # to choose only one node

# Define a function to programmatically select the root node:
# The function identifies cells with the specified lineage.
# It finds the principal graph node most frequently associated with these cells.
# This node is used as the root for ordering cells in pseudotime.
get_earliest_principal_node <- function(cds, predicted.id = lineage_value){
  cell_ids <- which(colData(cds)[, "predicted.id"] == lineage_value)

  closest_vertex <-
  cds@principal_graph_aux[["UMAP"]]$pr_graph_cell_proj_closest_vertex
  closest_vertex <- as.matrix(closest_vertex[colnames(cds), ])
  root_pr_nodes <-
    igraph::V(principal_graph(cds)[["UMAP"]])$name[as.numeric(names
   (which.max(table(closest_vertex[cell_ids,]))))]

  root_pr_nodes
}

# Order cells in the trajectory with the selected root node.
cds <- order_cells(cds, root_pr_nodes=get_earliest_principal_node(cds))
```

Extract the pseudotime values and add to the Seurat object:

```{r extract-pseudotime, fig.width = 15, fig.height = 5, fig.fullwidth = TRUE, echo=TRUE}
seurat_obj <- AddMetaData(object = seurat_obj,
                          metadata = lineage1.cds@principal_graph_aux@listData$UMAP$pseudotime,
                          col.name = glue::glue("{lineage1_value}_pseudotime"))
 
seurat_obj <- AddMetaData(object = seurat_obj,
                          metadata = lineage2.cds@principal_graph_aux@listData$UMAP$pseudotime,
                          col.name = glue::glue("{lineage2_value}_pseudotime"))
seurat_obj <- AddMetaData(object = seurat_obj,
                          metadata = cds@principal_graph_aux@listData$UMAP$pseudotime,
                          col.name = glue::glue("{lineage_value}_pseudotime_programmatically"))


name <- paste0(plots_dir, "/", glue::glue("Plot-pseudotime.png"), sep = "")
#print(FeaturePlot(seurat_obj, c(lineage1_value, lineage2_value), pt.size = 0.1) & scale_color_viridis_c())
# Create the three FeaturePlots
p1 <- FeaturePlot(seurat_obj, glue::glue("{lineage1_value}_pseudotime"), pt.size = 0.1) + scale_color_viridis_c()
p2 <- FeaturePlot(seurat_obj, glue::glue("{lineage2_value}_pseudotime"), pt.size = 0.1) + scale_color_viridis_c()
p3 <- FeaturePlot(seurat_obj, glue::glue("{lineage_value}_pseudotime_programmatically"), pt.size = 0.1) + scale_color_viridis_c()

# Combine them in a single row
combined_plot <- p1 | p2 | p3  # patchwork syntax for horizontal layout
print(combined_plot)
ggsave(file = name, width = 15, height = 5, device = "png")
```

# Save output files

```{r save_seurat, echo=TRUE}
# Save metadata
metadata <- as_data_frame_seurat(seurat_obj, metadata = TRUE)
write_tsv(metadata, file = paste0(results_dir, "/", glue::glue("metadata.tsv"))) # Save metadata
  
# Save object
saveRDS(seurat_obj, file = paste0(results_dir, "/", "seurat_obj_pseudotime.rds"))
```

```{r echo=FALSE}
detach(params)
```

\pagebreak

# Session Info

```{r echo=FALSE}
sessionInfo()
```
