---
title: "Integrating with scRNA-seq data for sc-ATAC-Seq Analysis in 10X Genomics data"
author: "Antonia Chroni for SJCRH DNB_BINF_Core"
papersize: a4
fontsize: 11pt
links-as-notes: true
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    df_print: paged
    code_folding: hide
    toc_depth: 2
    highlight: tango
    number_sections: TRUE
  pdf_document:
    toc: TRUE
    highlight: tango
    number_sections: TRUE
    latex_engine: lualatex
    keep_tex: FALSE
    fig_caption: yes
    fig_crop: no
    fig_height: 2
    fig_width: 3
    toc_depth: 2
always_allow_html: TRUE
urlcolor: blue
linkcolor: black
citecolor: blue
geometry: margin=1in
header-includes: 
  - \usepackage{titling}
  - \usepackage{fancyhdr}
  - \usepackage{graphicx}
  - \usepackage{float}
params:
  data_file: '.'
  input_data: '.'
  reduction_value: '.' 
  condition_value1: '.'
  condition_value2: '.'
  condition_value3: '.'
  assay: '.'
  nfeatures_value: '.'
  ct_palette_file: '.'
  reference_dir: '.'
  reference_file_name: '.'
  celltype_reference: '.'
  group_by_variable: '.'
  root_dir: './'
  PROJECT_NAME: '.'
  PI_NAME: '.'
  TASK_ID: '.'
  PROJECT_LEAD_NAME: '.'
  DEPARTMENT: '.'
  LEAD_ANALYSTS: '.'
  GROUP_LEAD: '.'
  CONTACT_EMAIL: '.'
  PIPELINE: '.'
  START_DATE: '.'
  COMPLETION_DATE: '.'
---
```{r logo-file, echo=FALSE}
attach(params)
# Insert logo on the top of the html report 
logo_file <- file.path(root_dir, "figures", "img", "DNB-BINF-Core-logo.png")
htmltools::img(src = knitr::image_uri(logo_file), alt = "logo", style = "position:absolute; top:0; left:0; padding:0px; height:120px;")
detach(params)
```

\addtolength{\headheight}{2.0cm} 
\fancypagestyle{plain}{} 
\thispagestyle{fancy}
\fancyhead[L]{\includegraphics[height=120px]{`r logo_file`}}
\renewcommand{\headrulewidth}{0pt}

<style type="text/css">
:root {--DNB_BINF_Core_color: #00427B;}

h1.title {margin-top: 130px;
          margin-bottom: 25px;
          font-size: 36px;}

.nobullet li {list-style-type: none;}

.reporthead {font-size: 20px;}

body { /* Normal */
  font-size: 16px;
  font-style: Arial, Helvetica, sans-serif;}

h1 {color: var(--DNB_BINF_Core_color);
    font-size: 28px;
    margin-top: 50px;}

h2 {color: var(--DNB_BINF_Core_color);
    font-size: 20px;}

.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
  background-color: var(--DNB_BINF_Core_color);}
</style>

<a href="https://wiki.stjude.org/display/CAB">

</a>

\pagebreak

<div class="reporthead"><br/>
**PI: `r params$PI_NAME`**  
**Project: `r params$PROJECT_NAME`**  
Task: `r params$TASK_ID`  
Project Lead(s): `r params$PROJECT_LEAD_NAME`  
Department: `r params$DEPARTMENT`  

<br />  

DNB Bioinformatics Core Analysis Team: 
<br />  

>**Lead Analyst(s): `r params$LEAD_ANALYSTS`**  
>Group Lead: `r params$GROUP_LEAD`  
<br />
>**Contact E-mail:** `r params$CONTACT_EMAIL`  
>**DNB Bioinformatics Core Pipeline:** `r params$PIPELINE`  

Date started: `r params$START_DATE`  
Date completed:  `r params$COMPLETION_DATE`  
Report generated: `r format(Sys.time(), '%H:%M:%S %Z %m/%d/%Y')` \

Reviewed by: _____________________   Date: ____________ \
</div>
\pagebreak
  
# Information about this notebook

Here, we will use a data transfer method in the context of scATAC-seq to:

- Classify cells measured with scATAC-seq based on clustering results from scRNA-seq
- Co-embed scATAC-seq and scRNA-seq data

Overall the following integration procedure consists of the following steps:

- Identify ‘anchors’ between the ATAC-seq and RNA-seq datasets
- Transfer data between datasets (either transfer labels for classification, or impute RNA levels in the ATAC-seq data to enable co-embedding)

For more information, see [Integrating with scRNA-seq data](https://stuartlab.org/signac/articles/mouse_brain_vignette#integrating-with-scrna-seq-data) and [PBMC scATAC-seq Vignette](https://satijalab.org/seurat/archive/v3.0/atacseq_integration_vignette.html).

# Set up
```{r load-library, echo=TRUE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(Seurat)
  library(Signac)
  library(scooter)
  library(knitr)
  library(patchwork)
}) 
```

# Directories and paths to file Inputs/Outputs

```{r set-dir-and-file-names, echo=TRUE}
attach(params)
analysis_dir <- file.path(root_dir, "analyses", "integration-with-scrna-seq-data") 

# Input files
cell_types_palette_file <- file.path(root_dir, "figures", "palettes", ct_palette_file)
reference_file <- file.path(reference_dir, reference_file_name)

# Create results_dir
module_results_dir <- file.path(analysis_dir, "results")
if (!dir.exists(module_results_dir)) {
  dir.create(module_results_dir)}

results_dir <- file.path(module_results_dir, "01_integration_with_scrna_seq_data")
if (!dir.exists(results_dir)) {
  dir.create(results_dir)}

# Create plots directory
module_plots_dir <- file.path(analysis_dir, "plots") 
plots_dir <- file.path(module_plots_dir, "01_integration_with_scrna_seq_data") 
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)}

source(paste0(root_dir, "/figures/scripts/theme_plot.R"))
source(paste0(analysis_dir, "/util/function-cell-type-fractions.R"))
source(paste0(analysis_dir, "/util/co-embedding-cells.R"))
```

```{r echo=FALSE, warning=FALSE}
opts_chunk$set(fig.align='center',
               external=TRUE,
               echo=FALSE,
               warning=FALSE,
               fig.pos='H')
a4width <- 8.3
a4height <- 11.7
```

# Read seurat object

First, we load the scATAC-seq query object, which contains a pre-computed gene activity matrix from the `input_data` module. We set the default assay to `"RNA"` so that gene activity values can be directly compared with the scRNA-seq reference. As a preprocessing step, we also identify the most variable genes, which will be used for integration.

```{r read-object, echo=TRUE}
atac_query <- readRDS(data_file)
DefaultAssay(atac_query) <- 'GeneActivity'
atac_query <- FindVariableFeatures(object = atac_query, nfeatures = nfeatures_value)
```

# Integrating with scRNA-seq data 

To help interpret the scATAC-seq data, we can classify cells based on an scRNA-seq experiment from the same biological system. We utilize methods for cross-modality integration and label transfer, described [here](https://www.sciencedirect.com/science/article/pii/S0092867419305598?via%3Dihub).

```{r define-parameters-for-plots, echo=TRUE}
# Read color palette
palette_df <- readr::read_tsv(cell_types_palette_file, guess_max = 100000, show_col_types = FALSE) %>%
  mutate(cell_type_names = case_when(cell_type_names == "na_color" ~ "unassigned",
                                       TRUE ~ cell_type_names))

# Define and order palette
palette <- palette_df$hex_codes 
names(palette) <- palette_df$cell_type_names 
```

```{r read-file-with-reference, fig.width = 18, fig.height = 6, fig.fullwidth = TRUE, echo=TRUE}
# Load the pre-processed scRNA-seq data
rna_reference <- readRDS(reference_file)

# Rename column containing cell types in the reference
rna_reference@meta.data$celltype <- rna_reference@meta.data[[celltype_reference]]
print(table(rna_reference@meta.data$celltype))

# Plot
name <- paste0(plots_dir, "/", glue::glue("plot-scATAC-vs-scRNA-seq-cells.png"))
p1.atac <- DimPlot(atac_query, reduction = "umap") + NoLegend() + ggtitle("scATAC-seq")
p2.rna <- DimPlot(rna_reference, reduction = reduction_value, group.by = 'celltype', label = TRUE, repel = TRUE, label.size = 2.5, cols = palette) + ggtitle("scRNA-Seq") 
print(p1.atac + p2.rna)
ggsave(file = name, width = 18, height = 6, device = "png")
```

Now, we can identify anchors between the scATAC-seq dataset and the scRNA-seq dataset and use these anchors to transfer the celltype labels we learned from the scRNA-seq data to the scATAC-seq cells.

To transfer the cluster ids, we provide a vector of previously annotated cell type labels for the RNA to the refdata parameter. The output will contain a matrix with predictions and confidence scores for each ATAC-seq cell.

We can then examine the distribution of prediction scores and optionally filter out those cells with low scores. If most scores are above 0.5–0.7, that's generally a good sign.

```{r transfer-anchors, fig.width = 8, fig.height = 6, fig.fullwidth = TRUE, echo=TRUE}
cat("Find anchors\n")
transfer.anchors <- FindTransferAnchors(reference = rna_reference,
                                          query = atac_query, 
                                          reduction = "cca",
                                          dims = 1:30)
  
cat("Transferring predicted labels...\n")
predicted.labels <- TransferData(anchorset = transfer.anchors,
                                 refdata = rna_reference$celltype,
                                 weight.reduction = atac_query[['lsi']],
                                 dims = 2:30)
  
cat("Add predicted.labels to query dataset\n")
atac_query <- AddMetaData(atac_query, metadata = predicted.labels) # Add metadata
  
# Add metadata
new.data <- atac_query@meta.data %>%
  as.data.frame() %>%
  add_column(project = "project")

atac_query <- AddMetaData(atac_query, metadata = new.data)

# Create table
summary_table <- table(predicted.id = atac_query@meta.data$predicted.id) %>% # all labels
  as.data.frame() %>% 
  mutate(cells_number_all = Freq) %>% 
  dplyr::select(-Freq) 
  
# Plot histogram of prediction scores
name <- paste0(plots_dir, "/", glue::glue("plot-hist-prediction.score.max.png"))
print(hist(atac_query$prediction.score.max))
plot <- ggplot(atac_query@meta.data, aes(x = prediction.score.max)) +
  geom_histogram(binwidth = 0.05, fill = "skyblue", color = "white") +
  geom_vline(xintercept = 0.5, color = "red", linetype = "dashed", size = 1) +
  labs(title = "Histogram of Prediction Scores", x = "prediction.score.max", y = "Count") +
  theme_minimal()
print(plot)
ggsave(file = name, plot = plot, width = 8, height = 6, device = "png")

# Print thresholding summary
# How many cells were confidently labeled (e.g., with a score > 0.5)?
print(table(atac_query$prediction.score.max > 0.5))
```

## Number of cells for the scATAC-seq cell predictions

We will check the number of cells by summarizing the identified scATAC-seq cell predictions.

```{r, fig.align = "left", results = "asis", message = FALSE, warning = FALSE, echo = FALSE}
tables1 <- summary_table
cat("  \n<div align=\"center\" style=\"font-size:80%\">  \n")
print(knitr::kable(tables1, align = "lcccc", caption = "scATAC-seq cell predictions"))
cat("  \n</div>  \n")
cat("  \n\\pagebreak  \n")
```

## Number of cells for the scATAC-seq cell predictions per ID

We will check the number of cells by summarizing the identified scATAC-seq cell predictions per ID.

```{r, fig.align = "left", results = "asis", message = FALSE, warning = FALSE, echo = FALSE}
tables1 <- atac_query@meta.data %>% 
  as.data.frame() %>% 
  dplyr::count(predicted.id, ID) %>%
  pivot_wider(names_from = ID, values_from = n, values_fill = list(n = 0))
cat("  \n<div align=\"center\" style=\"font-size:80%\">  \n")
print(knitr::kable(tables1, align = "lcccc", caption = "scATAC-seq cell predictions per ID"))
cat("  \n</div>  \n")
cat("  \n\\pagebreak  \n")
```

# Identify number of cells with low-quality or no assignments of cell type annotations

```{r echo=TRUE}
atac_query@meta.data <- atac_query@meta.data %>%
  mutate(predicted.id = if_else(is.na(predicted.id), "unassigned", as.character(predicted.id)),
         predicted.id = if_else(predicted.id == "", "unassigned", predicted.id))

# unassigned cells 
cells_unassigned <- length(which(atac_query@meta.data$predicted.id == "unassigned"))
```

We will identify and rename the cells with low-quality or no assignments to `unassigned` (instead of NAs or empty strings). There were identified `r cells_unassigned` cells due to low-quality or no assignments of cell type annotations. We will include these cells in the following plots.

# Plotting scATAC-seq cell predictions

## Predicted cell types: scATAC-seq cells vs scRNA-seq cells

We can then visualize the predicted cell types on a UMAP representation of the scATAC-seq data. Ideally, the transferred labels should align well with the structure captured by the UMAP.

```{r plot-comparison, fig.width = 18, fig.height = 6, fig.fullwidth = TRUE, echo=TRUE}
name <- paste0(plots_dir, "/", glue::glue("plot-scATAC-seq-cells-vs-scRNA-seq-cells.png"))
#atac_query.filtered <- subset(atac_query, subset = prediction.score.max > 0.5)
#atac_query.filtered$predicted.id <- factor(atac_query.filtered$predicted.id, levels = levels(rna_reference))  # to make the colors match
# head(atac_query.filtered$predicted.id)
# You should see cell type names. If it's all NA, label transfer didn't assign identities properly.
#p1 <- DimPlot(atac_query.filtered, group.by = 'predicted.id', label = TRUE, repel = TRUE, label.size = 2.5, cols = palette) + ggtitle("scATAC-seq cells")
p1 <- DimPlot(atac_query, group.by = 'predicted.id', label = TRUE, repel = TRUE, label.size = 2.5, cols = palette) + ggtitle("scATAC-seq cells")
print(p1 + p2.rna)
ggsave(file = name, width = 18, height = 6, device = "png")
```

```{r plot-cell-types-project-fractions, fig.width = 6, fig.height = 5, fig.fullwidth = TRUE, echo=TRUE}
name <- paste0(plots_dir, "/", glue::glue("plot-scRNA-seq-project-fractions.png"))
p.project <- print(cell_type_fractions_cell_type_reference(df = atac_query@meta.data,
                                                           condition_plot = "project",
                                                           color_df = palette,
                                                           title_value = "scATAC-seq"))
ggsave(file = name, width = 6, height = 5, device = "png")
```

## scATAC-seq cells: Predicted cell types and fractions per ID

```{r}
# Get the number of unique values in the 'ID' column
unique_ids <- print(length(unique(atac_query@meta.data$ID)))

# Set the maximum number of plots per row (e.g., 3 plots per row)
max_plots_per_row <- 3

# Calculate the number of columns needed based on unique IDs
# Ensure that ncol is at least 1 and limit by max_plots_per_row
ncol_plots <- max(1, min(max_plots_per_row, unique_ids))
```

```{r plot-cell-types-ID, fig.width = 25, fig.height = 10, fig.fullwidth = TRUE, echo=TRUE}
name <- paste0(plots_dir, "/", glue::glue("plot-scATAC-seq-ID.png"))
p1.after <- print(DimPlot(atac_query, reduction = reduction_value, split.by = "ID", group.by = 'predicted.id', label = TRUE, repel = TRUE, label.size = 2.5, cols = palette) + 
                    ggtitle("scATAC-seq cells: Predicted cell types") +
                    facet_wrap(~ ID, ncol = ncol_plots, scales = "free")) # Limit the number of columns per row
ggsave(file = name, width = 25, height = 10, device = "png")
```

```{r plot-cell-types-ID-fractions, fig.width = 6, fig.height = 5, fig.fullwidth = TRUE, echo=TRUE}
name <- paste0(plots_dir, "/", glue::glue("plot-scATAC-seq-ID-fractions.png"))
p.ID <- print(cell_type_fractions_cell_type_reference(df = atac_query@meta.data,
                                                      condition_plot = "ID",
                                                      color_df = palette,
                                                      title_value = "scATAC-seq"))
ggsave(file = name, width = 6, height = 5, device = "png")
```

## scATAC-seq cells: Predicted cell types and fractions per `r condition_value1`

```{r}
# Get the number of unique values in the 'ID' column
unique_ids <- print(length(unique(atac_query@meta.data[[condition_value1]])))

# Set the maximum number of plots per row (e.g., 3 plots per row)
max_plots_per_row <- 3

# Calculate the number of columns needed based on unique IDs
# Ensure that ncol is at least 1 and limit by max_plots_per_row
ncol_plots <- max(1, min(max_plots_per_row, unique_ids))
```

```{r plot-cell-types-condition1, fig.width = 18, fig.height = 6, fig.fullwidth = TRUE, echo=TRUE}
name <- paste0(plots_dir, "/", glue::glue("plot-scATAC-seq-condition1.png"))
p1.after <- print(DimPlot(atac_query, reduction = reduction_value, split.by = condition_value1, group.by = 'predicted.id', label = TRUE, repel = TRUE, label.size = 2.5, cols = palette) + 
                    ggtitle("scATAC-seq cells: Predicted cell types") +
                    facet_wrap(as.formula(paste("~", condition_value1)), ncol = ncol_plots, scales = "free")) # Limit the number of columns per row)
ggsave(file = name, width = 18, height = 6, device = "png")
```

```{r plot-cell-types-condition-fractions1, fig.width = 8, fig.height = 6, fig.fullwidth = TRUE, echo=TRUE}
name <- paste0(plots_dir, "/", glue::glue("plot-scATAC-seq-condition-fractions1.png"))
p.condition <- print(cell_type_fractions_cell_type_reference(df = atac_query@meta.data,
                                         condition_plot = condition_value1,
                                         color_df = palette,
                                         title_value = "scATAC-seq"))
ggsave(file = name, width = 8, height = 6, device = "png")
```

## scATAC-seq cells: Predicted cell types and fractions per `r condition_value2`

```{r plot-cell-types-condition2, fig.width = 18, fig.height = 6, fig.fullwidth = TRUE, echo=TRUE}
if (!is.null(condition_value2)) {
  
  # Get the number of unique values in the 'ID' column
  unique_ids <- print(length(unique(atac_query@meta.data[[condition_value2]])))

  # Set the maximum number of plots per row (e.g., 3 plots per row)
  max_plots_per_row <- 3

  # Calculate the number of columns needed based on unique IDs
  # Ensure that ncol is at least 1 and limit by max_plots_per_row
  ncol_plots <- max(1, min(max_plots_per_row, unique_ids))
  
  name <- paste0(plots_dir, "/", glue::glue("plot-scATAC-seq-condition2.png"))
  p1.after <- print(DimPlot(atac_query, reduction = reduction_value, split.by = condition_value2, group.by = 'predicted.id', label = TRUE, repel = TRUE, label.size = 2.5, cols = palette) + 
                    ggtitle("scATAC-seq cells: Predicted cell types") +
                    facet_wrap(as.formula(paste("~", condition_value2)), ncol = ncol_plots, scales = "free")) # Limit the number of columns per row)
  ggsave(file = name, width = 18, height = 6, device = "png")
}
```

```{r plot-cell-types-condition-fractions2, fig.width = 8, fig.height = 6, fig.fullwidth = TRUE, echo=TRUE}
if (!is.null(condition_value2)) {
  name <- paste0(plots_dir, "/", glue::glue("plot-scATAC-seq-condition-fractions2.png"))
  p.condition <- print(cell_type_fractions_cell_type_reference(df = atac_query@meta.data,
                                                                  condition_plot = condition_value2,
                                                                  color_df = palette,
                                                                  title_value = "scATAC-seq"))
  ggsave(file = name, width = 8, height = 6, device = "png")
}
```

## scATAC-seq cells: Predicted cell types and fractions per `r condition_value3`

```{r plot-cell-types-condition3, fig.width = 18, fig.height = 6, fig.fullwidth = TRUE, echo=TRUE}
if (!is.null(condition_value3)) {

  # Get the number of unique values in the 'ID' column
  unique_ids <- print(length(unique(atac_query@meta.data[[condition_value3]])))

  # Set the maximum number of plots per row (e.g., 3 plots per row)
  max_plots_per_row <- 3

  # Calculate the number of columns needed based on unique IDs
  # Ensure that ncol is at least 1 and limit by max_plots_per_row
  ncol_plots <- max(1, min(max_plots_per_row, unique_ids))
  
  name <- paste0(plots_dir, "/", glue::glue("plot-scATAC-seq-condition3.png"))
  p1.after <- print(DimPlot(atac_query, reduction = reduction_value, split.by = condition_value3, group.by = 'predicted.id', label = TRUE, repel = TRUE, label.size = 2.5, cols = palette) + 
                    ggtitle("scATAC-seq cells: Predicted cell types") +
                    facet_wrap(as.formula(paste("~", condition_value3)), ncol = ncol_plots, scales = "free")) # Limit the number of columns per row)
  ggsave(file = name, width = 18, height = 6, device = "png")
}
```

```{r plot-cell-types-condition-fractions3, fig.width = 8, fig.height = 6, fig.fullwidth = TRUE, echo=TRUE}
if (!is.null(condition_value3)) {
  name <- paste0(plots_dir, "/", glue::glue("plot-scATAC-seq-condition-fractions3.png"))
  p.condition <- print(cell_type_fractions_cell_type_reference(df = atac_query@meta.data,
                                         condition_plot = condition_value3,
                                         color_df = palette,
                                         title_value = "scATAC-seq"))
  ggsave(file = name, width = 8, height = 6, device = "png")
}
```

# Co-embedding

Finally, to visualize all the cells together, we can co-embed the scRNA-seq and scATAC-seq cells in the same low dimensional space. Here, we use the same anchors used earlier to transfer cell type labels to impute RNA-seq values for the scATAC-seq cells. We then merge the measured and imputed scRNA-seq data and run a standard UMAP analysis to visualize all the cells together. Note that this step is for visualization purposes only and is not a necessary part of the data transfer analysis.

Here we will also plot all cells colored by either their assigned cell type or their predicted cell type from the data transfer procedure. Upon inspection of the UMAP embeddings, we may notice clusters of cells that appear to be unique to a single modality. For example, certain cell populations might be detected exclusively in the transcriptomic dataset and not in the chromatin accessibility data. This likely reflects biological differences, such as cell types with minimal or absent nuclear material, or those in which chromatin accessibility is decoupled from transcriptional activity. As such, these cells may not align well across modalities and would be expected to appear only in one dataset.

```{r co-embedding, fig.width = 18, fig.height = 6, fig.fullwidth = TRUE, echo=TRUE}
coembed <- co_embedding_cells(reference_obj = rna_reference, seurat_obj.query = atac_query, transfer.anchors = transfer.anchors, group_by_variable = group_by_variable)
```

# Save output files

```{r save-output, echo=TRUE}
# Identify columns with a '.1' suffix
cols_to_remove <- grep("\\.1$", colnames(atac_query@meta.data), value = TRUE)
  
# Exclude columns that match the specific patterns (e.g., {assay}_snn_res.0.1, {assay}_snn_res.1, {assay}_snn_res.10)
cols_to_remove <- cols_to_remove[!grepl(glue::glue("^{assay}_snn_res\\.0\\.1$"), cols_to_remove) & 
                                     !grepl(glue::glue("^{assay}_snn_res\\.1$"), cols_to_remove) &
                                     !grepl(glue::glue("^{assay}_snn_res\\.10$"), cols_to_remove)]
  
# Remove the columns
atac_query@meta.data <- atac_query@meta.data[, !colnames(atac_query@meta.data) %in% cols_to_remove]
head(atac_query@meta.data)

metadata <- as_data_frame_seurat(atac_query, metadata = TRUE)
write_tsv(metadata, file = paste0(results_dir, "/", "metadata", ".tsv")) # Save metadata
saveRDS(atac_query, file = paste0(results_dir, "/", "seurat_obj_reference.rds"))
```

```{r echo=FALSE}
detach(params)
```

\pagebreak

# Session Info

```{r echo=FALSE}
sessionInfo()
```

