---
title: "Calling peaks per cell type analysis for sc-ATAC-Seq Analysis in 10X Genomics data"
author: "Antonia Chroni for SJCRH DNB_BINF_Core"
papersize: a4
fontsize: 11pt
links-as-notes: true
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    df_print: paged
    code_folding: hide
    toc_depth: 2
    highlight: tango
    number_sections: TRUE
  pdf_document:
    toc: TRUE
    highlight: tango
    number_sections: TRUE
    latex_engine: lualatex
    keep_tex: FALSE
    fig_caption: yes
    fig_crop: no
    fig_height: 2
    fig_width: 3
    toc_depth: 2
always_allow_html: TRUE
urlcolor: blue
linkcolor: black
citecolor: blue
geometry: margin=1in
header-includes: 
  - \usepackage{titling}
  - \usepackage{fancyhdr}
  - \usepackage{graphicx}
  - \usepackage{float}
params:
  assay: '.'
  resolution_list: '.'
  future_globals_value: '.'
  n_value: '.'
  genome_name: '.'
  ident.1_value: '.'
  condition: '.'
  min.cutoff_value: '.'
  OrgDb_value: '.'
  top_n_value_peaks: '.'
  root_dir: './'
  PROJECT_NAME: '.'
  PI_NAME: '.'
  TASK_ID: '.'
  PROJECT_LEAD_NAME: '.'
  DEPARTMENT: '.'
  LEAD_ANALYSTS: '.'
  GROUP_LEAD: '.'
  CONTACT_EMAIL: '.'
  PIPELINE: '.'
  START_DATE: '.'
  COMPLETION_DATE: '.'
---

```{r logo-file, echo=FALSE}
attach(params)
# Insert logo on the top of the html report 
logo_file <- file.path(root_dir, "figures", "img", "DNB-BINF-Core-logo.png")
htmltools::img(src = knitr::image_uri(logo_file), alt = "logo", style = "position:absolute; top:0; left:0; padding:0px; height:120px;")
detach(params)
```

\addtolength{\headheight}{2.0cm} 
\fancypagestyle{plain}{} 
\thispagestyle{fancy}
\fancyhead[L]{\includegraphics[height=120px]{`r logo_file`}}
\renewcommand{\headrulewidth}{0pt}

<style type="text/css">
:root {--DNB_BINF_Core_color: #00427B;}

h1.title {margin-top: 130px;
          margin-bottom: 25px;
          font-size: 36px;}

.nobullet li {list-style-type: none;}

.reporthead {font-size: 20px;}

body { /* Normal */
  font-size: 16px;
  font-style: Arial, Helvetica, sans-serif;}

h1 {color: var(--DNB_BINF_Core_color);
    font-size: 28px;
    margin-top: 50px;}

h2 {color: var(--DNB_BINF_Core_color);
    font-size: 20px;}

.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
  background-color: var(--DNB_BINF_Core_color);}
</style>

<a href="https://wiki.stjude.org/display/CAB">

</a>

\pagebreak

<div class="reporthead"><br/>
**PI: `r params$PI_NAME`**  
**Project: `r params$PROJECT_NAME`**  
Task: `r params$TASK_ID`  
Project Lead(s): `r params$PROJECT_LEAD_NAME`  
Department: `r params$DEPARTMENT`  

<br />  

DNB Bioinformatics Core Analysis Team: 
<br />  

>**Lead Analyst(s): `r params$LEAD_ANALYSTS`**  
>Group Lead: `r params$GROUP_LEAD`  
<br />
>**Contact E-mail:** `r params$CONTACT_EMAIL`  
>**DNB Bioinformatics Core Pipeline:** `r params$PIPELINE`  

Date started: `r params$START_DATE`  
Date completed:  `r params$COMPLETION_DATE`  
Report generated: `r format(Sys.time(), '%H:%M:%S %Z %m/%d/%Y')` \

Reviewed by: _____________________   Date: ____________ \
</div>
\pagebreak
  
# Information about this notebook

For more information, please see the [Calling peaks](https://stuartlab.org/signac/articles/peak_calling) vignette.

# Set up

```{r load-library, echo=TRUE}
attach(params)
suppressPackageStartupMessages({
  library(future)
  library(tidyverse)
  library(Seurat)
  library(Signac)
  library(knitr)
  library(DT)
  library(glue)

  library(htmlwidgets)
  library(clusterProfiler)

  library(enrichplot)

  library(org.Hs.eg.db)
  library(org.Mm.eg.db) # Mouse gene annotation (use org.Hs.eg.db for human)
  
  library(data.table)
  library(R.utils)
  library(Rsamtools)

  # Evaluate Seurat R expressions asynchronously when possible using future package
  options(future.globals.maxSize = future_globals_value, future.rng.onMisuse = "ignore") 
  plan(multisession, workers = parallelly::availableCores())
}) 
```

# Directories and paths to file Inputs/Outputs

```{r set-dir-and-file-names, echo=TRUE}
analysis_dir <- file.path(root_dir, "analyses", "integration-with-scrna-seq-data") 
data_dir <- file.path(root_dir, "analyses", "integration-with-scrna-seq-data", "results", "01_integration_with_scrna_seq_data")
module_results_dir <- file.path(analysis_dir, "results")
module_plots_dir <- file.path(analysis_dir, "plots")

util_dir <- file.path(root_dir, "analyses", "cluster-cell-calling") 

# Input files
data_file <- file.path(data_dir, "seurat_obj_reference.rds")

# Create results_dir
results_dir <- file.path(module_results_dir, "02_peak_calling")
if (!dir.exists(results_dir)) {
  dir.create(results_dir)}

# Create plots directory
plots_dir <- file.path(module_plots_dir, "02_peak_calling") 
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)}

coverage_plots_dir <- file.path(plots_dir, "coverage_plots") 
if (!dir.exists(coverage_plots_dir)) {
  dir.create(coverage_plots_dir)}

go_plots_dir <- file.path(plots_dir, "go_plots") 
if (!dir.exists(go_plots_dir)) {
  dir.create(go_plots_dir)}

# Source functions
source(paste0(util_dir, "/util/run-clusterwise-GO-enrichment.R"))
source(paste0(analysis_dir, "/util/function_process_fragments.R"))

```

```{r echo=FALSE, warning=FALSE}
opts_chunk$set(fig.align='center',
               external=TRUE,
               echo=FALSE,
               warning=FALSE,
               fig.pos='H')
a4width <- 8.3
a4height <- 11.7
```

# Read seurat object

First, we will use the seurat object as generated from the pipeline in the `integration-with-scrna-seq-data` module. 

```{r read-object, echo=TRUE}
seurat_obj <- readRDS(data_file)
DefaultAssay(seurat_obj) <- assay

Idents(seurat_obj) <- seurat_obj$predicted.id # Set Idents(seurat_obj) to match the group.by parameter you will use
table(Idents(seurat_obj)) # View cell identities, get summary table
```

```{r read-project-metadata, echo=TRUE}
metadata_dir <- yaml$metadata_dir
metadata_file <- yaml$metadata_file
project_metadata_file <- file.path(metadata_dir, metadata_file) # metadata input file

project_metadata <- read.csv(project_metadata_file, sep = "\t", header = TRUE)
sample_name <- unique(as.character(project_metadata$ID))
sample_name <- sort(sample_name, decreasing = FALSE)
print(sample_name)
```

```{r process-data, echo=TRUE}
# Example YAML
yaml_files <- list(metadata_dir = yaml$metadata_dir, metadata_file = yaml$metadata_file, data_dir = yaml$data_dir)

# Run
process_fragments(yaml_files)
```

```{r create-fragment-list, echo=TRUE}
# Build a list of fragment file paths, one for each sample
# These are the corrected .tsv.gz files where the barcodes were prefixed with the sample ID
fragment_paths <- file.path(yaml$data_dir, sample_name, "outs", "fragments.tsv.gz")

fragment_list <- mapply(function(sample, frag_path) {
  matching_cells <- grep(paste0("^", sample, ":"), colnames(seurat_obj), value = TRUE)

  CreateFragmentObject(path = frag_path, cells = matching_cells)
}, sample_name, fragment_paths, SIMPLIFY = FALSE)

# Assign the list of Fragment objects to the assay with the peaks
seurat_obj[[assay]]@fragments <- fragment_list
# lapply(Fragments(seurat_obj), function(x) x@path) # Optional: confirm it worked
```

# Calling Peaks (with MACS2) on grouped fragments by cell type

We use the [`CallPeaks()`](https://stuartlab.org/signac/reference/callpeaks) function from the **Signac** package to identify open chromatin regions (peaks) across different cell types or clusters from the raw fragment data. This function:

- Groups cells using a metadata column (here, `seurat_obj$predicted.id`)
- Calls peaks on the fragment data for each group
- Internally runs MACS2 to identify accessible chromatin regions


```{r call-peaks, echo=TRUE}
# Now call peaks on the Seurat object
peaks <- CallPeaks(
  object = seurat_obj,
  group.by = "predicted.id")

write.table(peaks, file = paste(results_dir, "/", "peaks_all.tsv", sep = ""), sep = "\t", quote = FALSE, row.names = FALSE)
```

## Interactive Table of Calling peaks by cell type

The results are returned as a GRanges object, with an additional metadata column listing the cell types that each peak was identified in:

```{r call-peaks-interactive-table, echo=TRUE}
# Interactive Table
datatable(as.data.frame(peaks), 
          options = list(pageLength = 5, 
                         autoWidth = TRUE, 
                         server = TRUE), 
          filter = "top")
```

## Visualizing Highly Variable Peaks for the `r assay` assay

Next, we scale the accessibility data for the selected features and generate a heatmap showing the accessibility profiles across all cells. Here, we select the top variable peaks (based on variability across cells), which can highlight regulatory elements that distinguish different cell states.

``` {r plot-peaks, fig.width = 25, fig.height = 15, fig.fullwidth = TRUE, echo=TRUE, include=FALSE}
seurat_obj <- ScaleData(seurat_obj)

# Select top 100 most variable peaks (or any subset)
top_peaks <- head(VariableFeatures(seurat_obj), 100)

name <- paste0(plots_dir, "/", glue::glue("Heatmap_peaks.png"), sep = "")
DoHeatmap(seurat_obj, features = top_peaks)  # or specify your own peaks
ggsave(file = name, width = 25, height = 15, device = "png")
```

### Heatmap Visualization of Selected Peaks

``` {r plot-peaks-subset, fig.width = 25, fig.height = 15, fig.fullwidth = TRUE, echo=TRUE}
name <- paste0(plots_dir, "/", glue::glue("Heatmap_peaks_subset.png"), sep = "")
print(DoHeatmap(subset(seurat_obj, downsample = 100), features = top_peaks))
ggsave(file = name, width = 25, height = 15, device = "png")
```


# Find Differentially Accessible Peaks between cell types 

We aim to identify and visualize the top differentially accessible peaks (DAPs) for each cell type. To achieve this, we use the [FindAllMarkers](https://satijalab.org/seurat/reference/findallmarkers) function, which performs differential accessibility analysis across the cell types defined in the predicted.id metadata column. This allows us to highlight peaks that are specifically enriched in accessibility within each cell type compared to the others.

```{r find-markers, echo=TRUE}
# Get unique cell types from predicted.id
cell_types <- unique(seurat_obj$predicted.id)

# Find markers 
# Try adjusting min.pct or logfc.threshold to be less stringent - useful for cases with groups of very few cells
markers <- FindAllMarkers(seurat_obj, group.by = "predicted.id", only.pos = TRUE,
                          min.pct = 0.1, logfc.threshold = 0.1,
                          min.cells.group = 1)
write.table(markers, file = paste(results_dir, "/", "markers_all.tsv", sep = ""), sep = "\t", quote = FALSE, row.names = FALSE)
```

```{r find-markers-interactive-table, echo=TRUE}
# Interactive Table
datatable(markers, 
          options = list(pageLength = 5, 
                         autoWidth = TRUE, 
                         server = TRUE), 
          filter = "top")
```

We will filter out and keep the first 3 top markers by cell type.

```{r}
top_markers <- markers %>% group_by(cluster) %>% top_n(n = 3, wt = avg_log2FC)
print(top_markers)
```

```{r top-markers-interactive-table, echo=TRUE}
# Interactive Table
datatable(top_markers, 
          options = list(pageLength = 5, 
                         autoWidth = TRUE, 
                         server = TRUE), 
          filter = "top")
```

## Plotting genomic regions across cell types

We will loop over each cell type and generate CoveragePlot by gene and marker. We can plot the frequency of Tn5 integration across regions of the genome for cells grouped by cluster, cell type, or any other metadata stored in the object for any genomic region using the [CoveragePlot()](https://stuartlab.org/signac/reference/coverageplot) function. These plots represent pseudo-bulk chromatin accessibility, where accessibility signal from all cells in a group is aggregated and visualized as a single track (thanks to Andrew Hill for giving the inspiration for this function in his excellent [blog post](https://andrewjohnhill.com/blog/2019/04/12/streamlining-scatac-seq-visualization-and-analysis/)). Alongside these accessibility tracks we can visualize other important information including gene annotation, peak coordinates, and genomic links (if they’re stored in the object). For more information, see [Plotting genomic regions](https://stuartlab.org/signac/articles/pbmc_vignette#plotting-genomic-regions) and [Visualization vignette](https://stuartlab.org/signac/articles/visualization).

```{r coverage-plot, fig.width = 10, fig.height = 6, fig.fullwidth = TRUE, echo=TRUE}
for (i in seq_len(nrow(top_markers))) {
  gene <- top_markers$gene[i]
  ct <- top_markers$cluster[i]
  
  message("Plotting ", gene, " for cell type ", ct)
  
  p <- CoveragePlot(
    object = seurat_obj,
    region = gene,
    group.by = "predicted.id",
    idents = ct,
    ranges = peaks,
    ranges.title = "MACS2",
    assay = assay) 
  
  print(p)
  
  # Save the plot as PNG
  plot_path <- file.path(coverage_plots_dir, paste0("CoveragePlot_", ct, "_", gene, ".png"))
  ggsave(plot_path, plot = p, width = 10, height = 6)
  
  message("Saved plot to: ", plot_path)
}
```

## Plotting shared genomic regions across cell types

```{r coverage-plot-shared-peaks, fig.width = 10, fig.height = 6, fig.fullwidth = TRUE, echo=TRUE}
# If `markers` contains DE peaks across cell types (from Signac peak assay)
shared_peaks <- markers %>%
  group_by(gene) %>%
  summarize(n_groups = n_distinct(cluster)) %>%
  filter(n_groups > 1) %>%
  arrange(desc(n_groups))

region_to_plot <- shared_peaks$gene[1]

# Step 2: Select top N shared peaks (e.g., top 3)
top_n_peaks <- head(shared_peaks$gene, 3)

# Step 3: Loop over each shared peak and generate a CoveragePlot
for (region_to_plot in top_n_peaks) {
  
  message("Plotting shared peak: ", region_to_plot)
  
  p <- CoveragePlot(
    object = seurat_obj,
    region = region_to_plot,
    group.by = "predicted.id",  # show all groups in same plot
    ranges = peaks,
    ranges.title = "MACS2",
    assay = assay  
  )
  
  print(p)
  # Save plot
  plot_path <- file.path(coverage_plots_dir, paste0("CoveragePlot_all_cell_types_shared_genes_", region_to_plot, ".png"))
  ggsave(plot_path, plot = p, width = 10, height = 6)
  
  message("Saved plot to: ", plot_path)
}
``` 

# Peak coordinates

Interpreting peak coordinates alone can be challenging. To gain functional insight, we use [ClosestFeature()](https://stuartlab.org/signac/reference/closestfeature) function to annotate each peak with its nearest gene.

```{r find-closest-gene-peaks, echo=TRUE}
closest_genes_list <- list()

for (clust in unique(markers$cluster)) {
  message("Processing cluster: ", clust)
  
  open_peaks <- markers %>%
    filter(cluster == clust, avg_log2FC > 1) %>%
    pull(gene)
  
  message("Number of open peaks before filtering: ", length(open_peaks))
  
  open_peaks <- open_peaks[open_peaks %in% rownames(seurat_obj[[assay]])]
  
  message("Number of open peaks after filtering: ", length(open_peaks))
  
  if (length(open_peaks) > 0) {
    closest <- ClosestFeature(seurat_obj, regions = open_peaks)
    closest <- closest %>% mutate(cluster = clust)
    closest_genes_list[[as.character(clust)]] <- closest
  } else {
    message("No open peaks to process for cluster: ", clust)
  }
}

# Combine all clusters into a single data frame and save to file
closest_genes_all <- do.call(rbind, closest_genes_list)
write.table(closest_genes_all, file = file.path(results_dir, "peak_called_in_wise_closest_genes.tsv"), sep = "\t", quote = FALSE, row.names = FALSE)
```

We can follow up these results by performing gene ontology (GO) enrichment analysis on the gene sets identified using [ClosestFeature()](https://stuartlab.org/signac/reference/closestfeature). Several R packages support this, including popular tools like GOstats and clusterProfiler.

# Gene Ontology (GO) enrichment analysis using clusterProfiler

In this step, we conduct Gene Ontology (GO) enrichment analysis to identify biological processes associated with cluster-specific accessible regions (e.g., differentially accessible peaks near genes). Using the [clusterProfiler](https://bioconductor.org/packages/devel/bioc/html/clusterProfiler.html) package, we assess whether any GO terms are significantly enriched among these gene sets. This approach helps interpret the functional implications of differential chromatin accessibility across clusters, i.e., predicted cell type.

```{r go-clusterProfiler, fig.width = 15, fig.height = 8, fig.fullwidth = TRUE, echo=TRUE}
ego_results <- run_clusterwise_GO_enrichment(closest_genes_all = closest_genes_all, OrgDb_value = OrgDb_value, plots_dir = go_plots_dir)
```

```{r echo=FALSE}
detach(params)
```

\pagebreak

# Session Info

```{r echo=FALSE}
sessionInfo()
```

