---
title: "Updates: `r PROJECT_NAME`"
author: "Antonia Chroni for SJCRH DNB_BINF_Core"
date: "`r Sys.Date()`"
papersize: a4
fontsize: 11pt
links-as-notes: true
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    df_print: paged
    code_folding: hide
    toc_depth: 2
    highlight: tango
    number_sections: TRUE
  pdf_document:
    toc: TRUE
    highlight: tango
    number_sections: TRUE
    latex_engine: lualatex
    keep_tex: FALSE
    fig_caption: yes
    fig_crop: no
    fig_height: 2
    fig_width: 3
    toc_depth: 2
always_allow_html: TRUE
urlcolor: blue
linkcolor: black
citecolor: blue
geometry: margin=1in
header-includes: 
  - \usepackage{titling}
  - \usepackage{fancyhdr}
  - \usepackage{graphicx}
  - \usepackage{float}
params:
  cellranger_parameters: '.'
  cancer_type: '.'
  cohort_value: '.'
  genome_reference: '.'
  root_dir: './'
  PROJECT_NAME: '.'
  PI_NAME: '.'
  TASK_ID: '.'
  PROJECT_LEAD_NAME: '.'
  DEPARTMENT: '.'
  LEAD_ANALYSTS: '.'
  GROUP_LEAD: '.'
  CONTACT_EMAIL: '.'
  PIPELINE: '.'
  START_DATE: '.'
  COMPLETION_DATE: '.'
---

```{r logo-file, echo=FALSE}
attach(params)
# Insert logo on the top of the html report 
logo_file <- file.path(root_dir, "figures", "img", "DNB-BINF-Core-logo.png")
htmltools::img(src = knitr::image_uri(logo_file), alt = "logo", style = "position:absolute; top:0; left:0; padding:0px; height:120px;")
detach(params)
```

\addtolength{\headheight}{2.0cm} 
\fancypagestyle{plain}{} 
\thispagestyle{fancy}
\fancyhead[L]{\includegraphics[height=120px]{`r logo_file`}}
\renewcommand{\headrulewidth}{0pt}

<style type="text/css">
:root {--DNB_BINF_Core_color: #00427B;}

h1.title {margin-top: 130px;
          margin-bottom: 25px;
          font-size: 36px;}

.nobullet li {list-style-type: none;}

.reporthead {font-size: 20px;}

body { /* Normal */
  font-size: 16px;
  font-style: Arial, Helvetica, sans-serif;}

h1 {color: var(--DNB_BINF_Core_color);
    font-size: 28px;
    margin-top: 50px;}

h2 {color: var(--DNB_BINF_Core_color);
    font-size: 20px;}

.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
  background-color: var(--DNB_BINF_Core_color);}
</style>

<a href="https://wiki.stjude.org/display/CAB">

</a>

\pagebreak

<div class="reporthead"><br/>
**PI: `r params$PI_NAME`**  
**Project: `r params$PROJECT_NAME`**  
Task: `r params$TASK_ID`  
Project Lead(s): `r params$PROJECT_LEAD_NAME`  
Department: `r params$DEPARTMENT`  

<br />  

DNB Bioinformatics Core Analysis Team: 
<br />  

>**Lead Analyst(s): `r params$LEAD_ANALYSTS`**  
>Group Lead: `r params$GROUP_LEAD`  
<br />
>**Contact E-mail:** `r params$CONTACT_EMAIL`  
>**DNB Bioinformatics Core Pipeline:** `r params$PIPELINE`  

Date started: `r params$START_DATE`  
Date completed:  `r params$COMPLETION_DATE`  
Report generated: `r format(Sys.time(), '%H:%M:%S %Z %m/%d/%Y')` \

Reviewed by: _____________________   Date: ____________ \
</div>
\pagebreak

# Set up

```{r load-library, echo = FALSE}
suppressPackageStartupMessages({
  library(tidyverse)  
  library(knitr)
  library(DT)
})
```

# Directories and paths to file Inputs/Outputs

```{r set-dir-and-file-names, echo = FALSE}
attach(params)
analysis_dir <- file.path(root_dir, "analyses", "upstream-analysis") 
module_plots_dir <- file.path(analysis_dir, "plots") 
module_results_dir <- file.path(analysis_dir, "results") 
fastqc_module_results_dir <- file.path(root_dir, "analyses", "fastqc-analysis", "results") 
cellranger_module_results_dir <- file.path(root_dir, "analyses", "cellranger-analysis", "results") 
#metadata_dir <- yaml$metadata_dir

fastqc_results_dir <- file.path(fastqc_module_results_dir, "02-multiqc-reports") 
cellranger_results_dir <- file.path(cellranger_module_results_dir, "03_cellranger_count_summary", cellranger_parameters) 
fastqc_overrepresented_results_file <- file.path(fastqc_results_dir, "fastqc_overrepresented_sequences_plot.txt")
scDblFinder_results_dir <- file.path(module_results_dir, "02_scDblFinder") 
Filter_object_results_dir <- file.path(module_results_dir, "03_Filter_object") 

if (file.exists(fastqc_overrepresented_results_file)) {
  # File exists – proceed to read or plot
  message("Loading overrepresented sequences plot data...")
  overrep_data <- read.delim(fastqc_overrepresented_results_file, header = TRUE)
  # (Insert your processing or plotting code here)
} else {
  # File does not exist – skip and inform
  message("Skipping overrepresented sequences plot: file not found.")
}

cellranger_results_file <- file.path(cellranger_results_dir, "QC_Summary_CellRanger_Report.tsv")
scDblFinder_results_file <- file.path(scDblFinder_results_dir, "doublets_metrics.tsv")
cells_table_file <- file.path(Filter_object_results_dir, "number_of_cells_per_sample.tsv")
#metadata_file <- file.path(metadata_dir, "project_metadata.tsv") # metadata input file

summary_samples_table_file <- file.path(metadata_dir, metadata_file) # summary_samples_table file
```


# Methods

## Data

### `r PROJECT_NAME`

```{r read-table, echo = FALSE}
summary_samples_table <- readr::read_tsv(summary_samples_table_file, guess_max = 100000, show_col_types = FALSE)
```

```{r, fig.align = "left", results = "asis", message = FALSE, warning = FALSE, echo = FALSE}
# Create the data frame
tables1 <- summary_samples_table
cat("  \n<div align=\"center\" style=\"font-size:80%\">  \n")
print(knitr::kable(tables1, align = "lcccc", caption = "Summary of samples in the cohort"))
cat("  \n</div>  \n")
cat("  \n\\pagebreak  \n")
```

  
# Results

All results can be found here:

- Cluster users can access data from: `r root_dir`/analyses/
- Mac users can access data from: `r mac_results_filepath`/`r cohort_value`/analyses/
- `r results_filepath`/`r cohort_value`/analyses/


We used FastQC (v0.11.9) and multiqc (v1.25) to perform quality control tool for all libraries across the project from the raw sequencing reads. Next, we aligned these reads using CellRanger (v8.0.1) (10x Genomics). 

## fastqc-analysis

FastQC metrics such as per-base sequence quality, GC content, and adapter contamination can indicate the quality of the sequencing data and contribute to determining if the data is suitable for downstream analyses. Further validation steps, such as alignment or feature counting, will also help confirm if the data is suitable for further processing and analysis.

For more information, please see: `r results_filepath`/`r cohort_value`/analyses/fastqc-analysis/results

### Results: `r PROJECT_NAME` cohort

```{r read-fastqc-analysis, echo = TRUE, results = 'asis', error = FALSE}
#multiqc_general_stats_df <- read.table(multiqc_general_stats_results_file, header = TRUE, sep = "\t") %>%
#  arrange(Sample)

#multiqc_fastqc_df <- read.table(multiqc_fastqc_results_file, header = TRUE, sep = "\t") %>% 
#  mutate_if(is.numeric, ~ scales::comma(.)) %>%
#  arrange(Sample)

# Define the path to the overrepresented sequences results
fastqc_overrepresented_results_file <- file.path(fastqc_results_dir, "fastqc_overrepresented_sequences_plot.txt")

# Check if the file exists and read it if available
if (file.exists(fastqc_overrepresented_results_file)) {
  
  fastqc_overrepresented_df <- read.table(fastqc_overrepresented_results_file, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>%
    dplyr::rename_with(~ gsub(" ", "_", .)) %>%
    dplyr::mutate(Overrepresented_sequences_by_sample_pct = Top.overrepresented.sequence + Sum.of.remaining.overrepresented.sequences) %>%
    dplyr::mutate(across(where(is.numeric), ~ round(., 2))) %>%
    dplyr::arrange(Sample)
  } else {
    cat("**Overrepresented sequences data not found.** Skipping this section.\n")
}
```

#### Sequence Quality Histograms
- This graph indicates the overall quality scores for the sequences. If a large proportion of sequences have lower scores this can indicated a general quality issue with the sequencing run

#### Per Base Sequence Content
- Non-greyish bars (i.e. bars with sections of distinct colors) indicates a **Non-Uniform Distribution** with one or more nucleotides dominating at specific positions.
- Sharp spikes in specific nucleotides can indicate **adapter contamination** or other sources of bias.

#### Per Sequence GC Content
- If there is **no bell-shaped curve** (normal distribution), which can be caused by:
    - Bias in GC content.
    - Contamination.
    - Poor quality or truncated sequences.
    - Artifacts in library preparation.
    - Bias due to specific sequencing methods.

- In an ideal scenario, the **Per Sequence GC Content** plot should show a bell-shaped curve (normal distribution), where:
    - The majority of sequences will have GC content near the middle range (around 40–60% GC content), as most sequencing technologies tend to produce more even distributions in well-prepared, high-quality samples.
    - The edges of the plot (where GC content is very low or very high) should have few sequences.

#### Sequence Duplication Levels
- This visualization helps assess how many of your sequences are duplicates of others in your dataset.
- High levels of duplicates can be problematic and affect downstream analyses, such as differential expression or variant calling.
- It is important for evaluating the **quality** and **complexity** of your sequencing library.

#### Overrepresented Sequences by Sample
- For example, primer dimers of the template switch oligo **`AAGCAGTGGTATCAACGCAGAGTACATGGG`** should not exceed 5-10%.


```{r, fig.align = "left", results = "asis", message = FALSE, warning = FALSE, echo = FALSE, error = FALSE}
if (file.exists(fastqc_overrepresented_results_file)) {
  tables1 <- fastqc_overrepresented_df
  cat("  \n<div align=\"center\" style=\"font-size:80%\">  \n")
  print(knitr::kable(tables1, align = "lcccc", caption = "Overrepresented Sequences by Sample and Duplication Levels for Specific Samples"))
  cat("  \n</div>  \n")
  cat("  \n\\pagebreak  \n")
  } else {
    cat("**Overrepresented sequences data not found.** Skipping this section.\n")
} 
```

## cellranger-analysis

### Results: `r PROJECT_NAME` cohort

We aligned these reads to the `r genome_reference` genome.

For more information, please see: `r results_filepath`/`r cohort_value`/analyses/cellranger-analysis/results

```{r read-cellranger-analysis, echo = FALSE}
cellranger_results_df <- readr::read_tsv(cellranger_results_file, guess_max = 100000, show_col_types = FALSE) %>%
  rename_with(~ gsub(" ", "_", .)) %>% # Replace spaces with underscores in column names
  relocate(Sample_ID) %>%
  arrange(Sample_ID) %>%
  mutate_if(is.numeric, ~ scales::comma(.)) %>%
  select(Sample_ID, Estimated_number_of_cells, Fraction_of_all_fragments_in_cells, Confidently_mapped_read_pairs, Warnings, MajorWarnings) %>%
  mutate(Minor_Flags = Warnings,
         Major_Flags = MajorWarnings) %>%
  select(-Warnings, -MajorWarnings)  # Remove original columns to avoid redundancy
```

```{r, fig.align = "left", results = "asis", message = FALSE, warning = FALSE, echo = FALSE}
tables1 <- cellranger_results_df
cat("  \n<div align=\"center\" style=\"font-size:80%\">  \n")
print(knitr::kable(tables1, align = "lcccc", caption = "Summary of CellRanger QC per sample"))
cat("  \n</div>  \n")
cat("  \n\\pagebreak  \n")
```

## upstream-analysis

In this step, we will filter out low quality cells based on identification of ambient RNA, low quality cells, and/or doublets (some steps may be optional). 

### Step 1: Signac QC metrics

We applied the following filtering parameters, as specified in the `sc-epigenie` pipeline:

- pct_reads_in_peaks_value_upstream = `r pct_reads_in_peaks_value_upstream` (% of reads in peaks (signal-to-noise)	> 15 or > 20)
- blacklist_ratio_value_upstream = `r blacklist_ratio_value_upstream` (Fraction of reads in ENCODE blacklist regions (noise metric)	< 0.05))
- TSS.enrichment_value_upstream = `r TSS.enrichment_value_upstream` (Transcription start site enrichment score	> 2)
- nucleosome_signal_value_upstream = `r nucleosome_signal_value_upstream` (Nucleosome periodicity signal	< 4)
           
For more information, please see: `r results_filepath`/`r cohort_value`/analyses/upstream-analysis/plots/01_Signac_qc

For more information per sample, please see `./plots/01_Signac_qc/<SAMPLE>/Report-seurat-qc-<SAMPLE>-Sys.Date().html`.

### Step 2: Estimating and filtering out doublets

We estimated doublets for each library. Currently, we have not accounted for doublets in the filtering process, though this option is available for future use.

It is common in single cell experiments to have 10-20% of doublets, which makes accurate doublet detection critical. Here, the results appear reasonable with proportion of doublets being proportional to the number of cells captured.

For more information, please see: `r results_filepath`/`r cohort_value`/analyses/upstream-analysis/plots/02_scDblFinder

#### Results: `r PROJECT_NAME` cohort

```{r read-scDblFinder, echo = FALSE}
doublets_metrics <- readr::read_tsv(scDblFinder_results_file, guess_max = 100000, show_col_types = FALSE) 
```

```{r, fig.align = "left", results = "asis", message = FALSE, warning = FALSE, echo = FALSE}
tables1 <- doublets_metrics
cat("  \n<div align=\"center\" style=\"font-size:80%\">  \n")
print(knitr::kable(tables1, align = "lcccc", caption = "Summary of doublets detection per sample"))
cat("  \n</div>  \n")
cat("  \n\\pagebreak  \n")
```

### Step 3: Merging filtered data

Next, we merge count matrices from `step 1` after filtering out low quality cells via Seurat QC pipeline. Seurat object and metadata for the library along with UMAP embeddings are saved to be used for downstream analyses.

For more information, please see: `r results_filepath`/`r cohort_value`/analyses/upstream-analysis/plots/03_Filter_object

#### Results: `r PROJECT_NAME` cohort

```{r read-Filter-object, echo = FALSE}
cells_table <- readr::read_tsv(cells_table_file, guess_max = 100000, show_col_types = FALSE) 
```

```{r, fig.align = "left", results = "asis", message = FALSE, warning = FALSE, echo = FALSE}
tables1 <- cells_table
cat("  \n<div align=\"center\" style=\"font-size:80%\">  \n")
print(knitr::kable(tables1, align = "lcccc", caption = "Summary of number of cells per sample at each QC step"))
cat("  \n</div>  \n")
cat("  \n\\pagebreak  \n")
```

### Step 4: Final QC summary report

Lastly, we provide a final QC summary report containing graphs and summary tables across each QC step: `fastqc-analysis`, `cellranger-analysis` and  `upstream-analysis` modules.

For more information, please see: `r results_filepath`/`r cohort_value`/analyses/upstream-analysis/plots/04_Final_summary_report


## integrative-analysis

For more information, please see: `r results_filepath`/`r cohort_value`/analyses/integrative-analysis/plots
 
 - The single-cell sequencing dataset was integrated using the **`r integration_method`** algorithm. 


## cluster-cell-calling

For more information, please see: `r results_filepath`/`r cohort_value`/analyses/cluster-cell-calling/plots

   - We used the **`r resolution_clustering_module`** resolution list for calling clusters. That included the following number of resolutions: **`r resolution_values`**.
   - We used the **`r resolution_list_find_markers`** resolution to identify differentially accessible peaks per cluster and explored the **`r top_n_value_peaks`** peaks.


# Future directions

- **QC upstream analyses**: 

- **Downstream analyses**: 

